A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-08-13 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: worked on some TODOs

The paper aims to ask the following question: can we, by arguing from necessity, find a minimal set of physical requirements for a state space of a physical system that allows us to directly connect to all the measure theoretic structures currently used in physics (i.e. probability measures, Liouville measure to count states, quasi-probability, ...) and point to a straightforward generalization of such structures? We will show that this is indeed possible.\footnote{Note that we believe that the requirements we set, though necessary, are not sufficient to generalize the Lie algebra given by the classical Poisson brackets and the quantum commutators, meaning they cannot fully recover mechanics. Nonetheless, they are enough to recover measure theoretic structures, the topic of this paper.}

@@

We will then review some results from the use of quasi-probability measures. The most used is the Wigner function,[cite] that represents each state as a signed probability measure over phase space, and expectations of observables are recovered through standard integration. While the Wigner function represents the most known instance, there are different choices depending on the problem at hand.[cite] The key insight is that it is not negative probability that characterizes quasi-probability,\footnote{Some representations guarantee the quasi-probability to be non-negative.} but the lack of mutual exclusivity of the points of the space. The quasi-probability measures, in fact, are not defined over the extreme points, but over a suitable set $K$ such that the space of signed probability measures is isomorphic as a vector space to the one that embeds the ensemble space. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

@@

Each physical theory satisfies these axioms, even though it may be represented very differently mathematically. Classical mechanics, for example, is typically described in terms of the geometry of phase space. Quantum mechanics is most often defined in terms of complex inner product spaces, but can be reframed in terms of quasi-probability distributions. Therefore, while the above requirements are needed for a physical theory, it does not necessarily mean that it is convenient to describe problems or make calculations in terms of those definitions. In fact, those requirements interact with each other leading to a rich mathematical structure that connects to many of the structures we use in mathematical physics, and our goal is mainly to see how the different representations are connected. In particular, here we want to examine the different ways that measure theory can be used to represent ensembles in any given physical theory.

@@

%Goal of this section: quasi-probability representations are allowed by the embedding in a topological vector space. Choice of representations are guided by what observables are interesting for a specific problem.

@@

%Thing to say: all these representations are useful because they preserve the linear structure, meaning that convex combinations of ensembles are convex combinations of quasi-probability; expectation values are linear. Check, but it should be true.

% 1 describe quasiprobability approaches; go through examples; point out that the quasi-probability can be negative; that one can recover expectations;

\textbf{Quasiprobability representations.} Another way to represent quantum ensembles is through the use of quasiprobability distributions, of which the Wigner function is probably the most widely known. The Wigner transform $W : \Ens \to L^1(\mathbb{R}^2, \mathbb{R}) $, defined as \cite[eq.~11.8-3a]{mandel1995}
\begin{equation}
	W(\rho) = W_{\rho} (x, p) = 2 \int_{-\infty}^{\infty} \langle x + y | \rho | x - y \rangle e^{-2 i p y / \hbar} \; \mathrm dy,
\end{equation}
takes a mixed state $\rho$ and returns a real function $W_{\rho}$ of phase space that, like a probability density integrates to one, but, unlike a probability density, can be negative. Recall that an affine combination is a linear combination where all coefficients sum to one but are not necessarily non-negative. Therefore, in general, quasi-probability distributions can be understood as affine combinations. Remarkably, expectation values of polynomials of position and momentum in symmetric ordering coincide with expectations over the Wigner function. For example
\begin{equation}
	\tr \left( \frac{1}{2}[XP+PX] \rho \right) = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} xp \; W_{\rho} (x, p) dx dp \; .
\end{equation}
Moreover, the transform is an affine map: $W(\lambda \rho_1 + \bar{\lambda} \rho_2 ) = \lambda W(\rho_1) + \bar{\lambda} W(\rho_2)$.

In some problems, especially in quantum optics, normal ordering and anti-normal ordering are more useful. The Glauber P [cite] representation and the Husimi Q [cite] representation work like the Wigner function except that the expectations of polynomials recover the respective order. In other circumstances, quasi-probability distributions over observables different from position and momentum are more useful. For example, the Kirkwood-Dirac quasiprobability is used to represent two-time correlation functions over observables defined at different times \cite{lostaglio2023}. Quasi-probability representations are also used over discrete elements that may not even correspond to eigenstates of a particular operator. For example, the SIC-POVM \cite{fuchs2004} represents quantum states as affine combinations of pure states $P_i$ that are ``equally distant'' from each other, see Fig. \ref{SIC}, which means that no two elements are orthogonal to each other. Each ensemble $\rho$ can then be represented as
\begin{equation}
	\rho = \sum_{i=1}^{d^2} \lambda_i P_i = \sum_{i=1}^{d^2} \left[ (d^2 + 1) p_i - \frac{1}{d} \right] P_i \; ,
\end{equation}
where $p_i = \tr \left(P_i \rho\right)$ and $d$ is the dimensionality of the Hilbert space.

It should be clear that there are many ways to represent quantum states via quasi-probability distributions, and that the choice of representation is a matter of convenience for the problem at hand. All representations, though, have the following features in common. First, we have a map $W : \Ens \to Q(Z,\mathbb{R})$ that given each mixed state $\rho \in \Ens$ returns the corresponding quasi-probability function $w(z) \in Q(Z,\mathbb{R})$, where $Q(Z,\mathbb{R}) \subset L^1(Z,\mathbb{R})$ is the function space used by the specific representation. The space $Z$, in fact, can change from representation to representation: it is $\mathbb{R}^{2n}$ for a Wigner function, $\mathbb{C}^n$ for the Husimi Q and $[1, d^2] \subset \mathbb{N}$ for the SIC-POVM. All representations preserve convex combinations: $W(\sum_i \lambda_i \rho_i) = \sum_i \lambda_i W(\rho_i)$. That is, a statistical mixture of quasi-probability distributions represents the statistical mixture of the corresponding states. Since the map is linear, and expectation values are linear, one can express expectation values as integrals or sums over the quasi-probability $w(z)$.

@@

\textbf{Not about negative probability.} The inclusion of negative probability is often advertised as the main difference between classical probability and quasi-probability \cite{adhikary2020}. Sometimes these negative probabilities are justified as arising from constraints on probability distributions defined on different contexts. This is a mischaracterization. For example, the Husimi Q representation provides us with a non-negative quasi-probability function. That is, the Husimi Q is a non-negative density distribution that integrates to one, yet it is not a classical probability distribution.\footnote{Since the Husimi Q is defined as $Q_{\rho} ( \alpha ) = \frac{1}{\pi} \langle \alpha | \rho | \alpha \rangle$, it is proportional to the probability of measuring a coherent (i.e. Gaussian) state $| \alpha \rangle$, which is non-negative. } Similarly, a SIC-POVM characterizes the Bloch ball by four pure states that form a tetrahedron inscribed in the ball. Since we can also inscribe the Bloch ball inside a tetrahedron, we can represent all states of a two-state quantum system by a convex combination of the vertices, leading to a non-negative quasi-probability distribution over four elements.

Conversely, we can use signed measures to represent classical distributions. Suppose we have a two state classical discrete space, which means the space of all possible ensembles is spanned by the vector $\begin{bmatrix} p & (1-p) \end{bmatrix}$. Alternatively, we can write it as $p\begin{bmatrix}1 & 0\end{bmatrix} + (1-p)\begin{bmatrix}0 & 1\end{bmatrix}$ which highlights that we are writing each possible probability distribution in terms of the two possible cases. But we could also express each probability distribution in terms of two other distributions, as in $\left(2p - \frac{1}{2}\right)\begin{bmatrix}\frac{3}{4} & \frac{1}{4}\end{bmatrix} + \left(\frac{3}{2} - 2p\right)\begin{bmatrix}\frac{1}{4} & \frac{3}{4}\end{bmatrix}$. TODO: check. The coefficients now span between $\begin{bmatrix}-\frac{1}{2}, \frac{3}{2}\end{bmatrix}$, can be negative, but still sum to one. That is, in terms of this new affine basis, we need to use a signed probability.

Since we can represent classical states with signed probability measures, and quantum states with non-negative probability measures, the difference between classical probability and quantum quasi-probability does not lie in the use of negative probability. Instead, the crucial difference is that in classical probability all points of the sample space, and therefore disjoint regions, represent mutually exclusive events, and all mutually exclusive events are represented by disjoint regions. Mutually exclusive events can be prepared and observed independently from the others (i.e. they are orthogonal in the sense that they maximize entropy increase during mixture); we can assign some probability to some region without having to assign it to another disjoint region.\footnote{The only constraint is that the probability over the whole space is one.} This is not the case for quasi-probability in quantum mechanics.

@@

TODO: remove all the - in quasi-probability

@@

\textbf{Ensemble space embedding and signed measures.} We saw that negative probability is a red herring, and mutual exclusivity is key. But how exactly does mutual exclusivity give us classical probability? Moreover, are quasi-probability representations specific to quantum mechanics, or are they more general? This is where the perspective from the ensemble space framework allows us to answer these questions directly.

@@

Now the second question: why does quantum mechanics allow quasi-probability representations? Let $Z$ be the space over which we are defining our quasi-probability distributions. Note that, in general, $Z$ is not the space of states. Wigner functions, for example, are defined over phase space, which does not represent quantum states. Let $M^{\pm}(Z)$ be the space of all bounded signed measures over $Z$. Since the linear combination of two bounded signed measures is a bounded signed measure, $M^{\pm}(Z)$ is a vector space. Now let $M_1^{\pm}(Z)$ be the space of signed probability measures. While it is not a subspace of $M^{\pm}(Z)$ as it does not include the zero measure, it is an affine subspace. Therefore, up to the choice of an arbitrary origin, it is a vector space.

Recall that an ensemble space must be a convex subset of a vector space. If we are able to represent ensembles as quasi-probability distributions over $Z$, then, $M_1^{\pm}(Z)$ will be isomorphic to the embedding vector space $V$ as topological vector spaces.\footnote{The question of whether such a $Z$ always exists is still open for us. A possible conjecture is that $Z$ could be constructed from the statistical quantities $F_i$ that fully identify each ensemble in $\Ens$.} In this case, any element $v\in V$ can be represented uniquely by a signed probability measure $\mu_{v}$, regardless of whether it is an actual ensemble or not. Since statistical quantities $F_i$ are affine functions, we will have $F_i(v) = \int_Z F_i(z) d\mu_{v}$. In other words, there is nothing special about quantum mechanics that allows quasi-probability representations. It is the fact that the ensemble space embeds into a vector space, a general fact for all theories, that allows such representations.

@@

Quasiprobability distributions are very useful in practice, particularly because one can construct them to fit particular problems. Once one has identified a set of observables that are the most natural ones to define a problem, one may find a representation that is specifically suited for the case. While the quasi-probability cannot be interpreted as probability, it still provides a useful tool to describe and understand the problem and give some intuition.

However, this practical convenience is also a drawback for finding a more general approach to probability in spaces beyond classical mechanics. Crucially, the space $Z$ over which the quasi-probability measures are defined is not unique, and it is not defined by the structure of the ensemble space and, as a consequence, cannot tell us which quasi-probability measures represent valid ensembles and which do not.

As a comparison, in the previous case all probability measures were given over a uniquely defined set of points, the extreme points of the ensemble space, but the probability measure may not be unique. In this case, the quasi-probability associated to each ensemble is unique, but the space over which the quasi-probability is defined is not.

@@

\textbf{Uniqueness of the fraction capacity.} From a conceptual standpoint, the fraction capacity has a major advantage: it has a clear physical meaning and it is unique. Like in Choquet theory and unlike for quasi-probability, the domain over which the measure is defined is fully determined by the system, by the ensemble space. On the other hand, unlike Choquet theory and like quasi-probability, once the space is fixed, there is one unique fraction capacity for each ensemble. Moreover, the setup is more general than Choquet theory (i.e. the ensemble space need not be compact) and stems from necessary physical requirements. An in depth study on fraction capacities, then, is guaranteed to be applicable to all physical theories, including those yet to be formulated.

>>>>

The paper aims to ask the following question: can we, by arguing from necessity, find a minimal set of physical requirements for a state space of a physical system that allows us to directly connect to all the measure theoretic structures currently used in physics (i.e. probability measures, Liouville measure to count states, quasiprobability, ...) and point to a straightforward generalization of such structures? We will show that this is indeed possible.\footnote{Note that we believe that the requirements we set, though necessary, are not sufficient to generalize the Lie algebra given by the classical Poisson brackets and the quantum commutators, meaning they cannot fully recover mechanics. Nonetheless, they are enough to recover measure theoretic structures, the topic of this paper.}

@@

We will then review some results from the use of quasiprobability measures. The most used is the Wigner function,[cite] that represents each state as a signed probability measure over phase space, and expectations of observables are recovered through standard integration. While the Wigner function represents the most known instance, there are different choices depending on the problem at hand.[cite] The key insight is that it is not negative probability that characterizes quasiprobability,\footnote{Some representations guarantee the quasiprobability to be non-negative.} but the lack of mutual exclusivity of the points of the space. The quasiprobability measures, in fact, are not defined over the extreme points, but over a suitable set $K$ such that the space of signed probability measures is isomorphic as a vector space to the one that embeds the ensemble space. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

@@

Each physical theory satisfies these axioms, even though it may be represented very differently mathematically. Classical mechanics, for example, is typically described in terms of the geometry of phase space. Quantum mechanics is most often defined in terms of complex inner product spaces, but can be reframed in terms of quasiprobability distributions. Therefore, while the above requirements are needed for a physical theory, it does not necessarily mean that it is convenient to describe problems or make calculations in terms of those definitions. In fact, those requirements interact with each other leading to a rich mathematical structure that connects to many of the structures we use in mathematical physics, and our goal is mainly to see how the different representations are connected. In particular, here we want to examine the different ways that measure theory can be used to represent ensembles in any given physical theory.

@@

%Goal of this section: quasiprobability representations are allowed by the embedding in a topological vector space. Choice of representations are guided by what observables are interesting for a specific problem.

@@

%Thing to say: all these representations are useful because they preserve the linear structure, meaning that convex combinations of ensembles are convex combinations of quasiprobability; expectation values are linear. Check, but it should be true.

% 1 describe quasiprobability approaches; go through examples; point out that the quasiprobability can be negative; that one can recover expectations;

\textbf{Quasiprobability representations.} Another way to represent quantum ensembles is through the use of quasiprobability distributions, of which the Wigner function is probably the most widely known. The Wigner transform $W : \Ens \to L^1(\mathbb{R}^2, \mathbb{R}) $, defined as \cite[eq.~11.8-3a]{mandel1995}
\begin{equation}
	W(\rho) = W_{\rho} (x, p) = 2 \int_{-\infty}^{\infty} \langle x + y | \rho | x - y \rangle e^{-2 i p y / \hbar} \; \mathrm dy,
\end{equation}
takes a mixed state $\rho$ and returns a real function $W_{\rho}$ of phase space that, like a probability density integrates to one, but, unlike a probability density, can be negative. Recall that an affine combination is a linear combination where all coefficients sum to one but are not necessarily non-negative. Therefore, in general, quasiprobability distributions can be understood as affine combinations. Remarkably, expectation values of polynomials of position and momentum in symmetric ordering coincide with expectations over the Wigner function. For example
\begin{equation}
	\tr \left( \frac{1}{2}[XP+PX] \rho \right) = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} xp \; W_{\rho} (x, p) dx dp \; .
\end{equation}
Moreover, the transform is an affine map: $W(\lambda \rho_1 + \bar{\lambda} \rho_2 ) = \lambda W(\rho_1) + \bar{\lambda} W(\rho_2)$.

In some problems, especially in quantum optics, normal ordering and anti-normal ordering are more useful. The Glauber P [cite] representation and the Husimi Q [cite] representation work like the Wigner function except that the expectations of polynomials recover the respective order. In other circumstances, quasiprobability distributions over observables different from position and momentum are more useful. For example, the Kirkwood-Dirac quasiprobability is used to represent two-time correlation functions over observables defined at different times \cite{lostaglio2023}. Quasiprobability representations are also used over discrete elements that may not even correspond to eigenstates of a particular operator. For example, the SIC-POVM \cite{fuchs2004} represents quantum states as affine combinations of pure states $P_i$ that are ``equally distant'' from each other, see Fig. \ref{SIC}, which means that no two elements are orthogonal to each other. Each ensemble $\rho$ can then be represented as
\begin{equation}
	\rho = \sum_{i=1}^{d^2} \lambda_i P_i = \sum_{i=1}^{d^2} \left[ (d^2 + 1) p_i - \frac{1}{d} \right] P_i \; ,
\end{equation}
where $p_i = \tr \left(P_i \rho\right)$ and $d$ is the dimensionality of the Hilbert space.

It should be clear that there are many ways to represent quantum states via quasiprobability distributions, and that the choice of representation is a matter of convenience for the problem at hand. All representations, though, have the following features in common. First, we have a map $W : \Ens \to Q(Z,\mathbb{R})$ that given each mixed state $\rho \in \Ens$ returns the corresponding quasiprobability function $w(z) \in Q(Z,\mathbb{R})$, where $Q(Z,\mathbb{R}) \subset L^1(Z,\mathbb{R})$ is the function space used by the specific representation. The space $Z$, in fact, can change from representation to representation: it is $\mathbb{R}^{2n}$ for a Wigner function, $\mathbb{C}^n$ for the Husimi Q and $[1, d^2] \subset \mathbb{N}$ for the SIC-POVM. All representations preserve convex combinations: $W(\sum_i \lambda_i \rho_i) = \sum_i \lambda_i W(\rho_i)$. That is, a statistical mixture of quasiprobability distributions represents the statistical mixture of the corresponding states. Since the map is linear, and expectation values are linear, one can express expectation values as integrals or sums over the quasiprobability $w(z)$.

@@

\textbf{Not about negative probability.} The inclusion of negative probability is often advertised as the main difference between classical probability and quasiprobability \cite{adhikary2020}. Sometimes these negative probabilities are justified as arising from constraints on probability distributions defined on different contexts. This is a mischaracterization. For example, the Husimi Q representation provides us with a non-negative quasiprobability function. That is, the Husimi Q is a non-negative density distribution that integrates to one, yet it is not a classical probability distribution.\footnote{Since the Husimi Q is defined as $Q_{\rho} ( \alpha ) = \frac{1}{\pi} \langle \alpha | \rho | \alpha \rangle$, it is proportional to the probability of measuring a coherent (i.e. Gaussian) state $| \alpha \rangle$, which is non-negative. } Similarly, a SIC-POVM characterizes the Bloch ball by four pure states that form a tetrahedron inscribed in the ball. Since we can also inscribe the Bloch ball inside a tetrahedron, we can represent all states of a two-state quantum system by a convex combination of the vertices, leading to a non-negative quasiprobability distribution over four elements.

Conversely, we can use signed measures to represent classical distributions. Suppose we have a two state classical discrete space, which means the space of all possible ensembles is spanned by the vector $\begin{bmatrix} p & (1-p) \end{bmatrix}$. Alternatively, we can write it as $p\begin{bmatrix}1 & 0\end{bmatrix} + (1-p)\begin{bmatrix}0 & 1\end{bmatrix}$ which highlights that we are writing each possible probability distribution in terms of the two possible cases. But we could also express each probability distribution in terms of two other distributions, as in $\left(2p - \frac{1}{2}\right)\begin{bmatrix}\frac{3}{4} & \frac{1}{4}\end{bmatrix} + \left(\frac{3}{2} - 2p\right)\begin{bmatrix}\frac{1}{4} & \frac{3}{4}\end{bmatrix}$. The coefficients now span between $\begin{bmatrix}-\frac{1}{2}, \frac{3}{2}\end{bmatrix}$, can be negative, but still sum to one. That is, in terms of this new affine basis, we need to use a signed probability.

Since we can represent classical states with signed probability measures, and quantum states with non-negative probability measures, the difference between classical probability and quantum quasiprobability does not lie in the use of negative probability. Instead, the crucial difference is that in classical probability all points of the sample space, and therefore disjoint regions, represent mutually exclusive events, and all mutually exclusive events are represented by disjoint regions. Mutually exclusive events can be prepared and observed independently from the others (i.e. they are orthogonal in the sense that they maximize entropy increase during mixture); we can assign some probability to some region without having to assign it to another disjoint region.\footnote{The only constraint is that the probability over the whole space is one.} This is not the case for quasiprobability in quantum mechanics.

@@

@@

\textbf{Ensemble space embedding and signed measures.} We saw that negative probability is a red herring, and mutual exclusivity is key. But how exactly does mutual exclusivity give us classical probability? Moreover, are quasiprobability representations specific to quantum mechanics, or are they more general? This is where the perspective from the ensemble space framework allows us to answer these questions directly.

@@

Now the second question: why does quantum mechanics allow quasiprobability representations? Let $Z$ be the space over which we are defining our quasiprobability distributions. Note that, in general, $Z$ is not the space of states. Wigner functions, for example, are defined over phase space, which does not represent quantum states. Let $M^{\pm}(Z)$ be the space of all bounded signed measures over $Z$. Since the linear combination of two bounded signed measures is a bounded signed measure, $M^{\pm}(Z)$ is a vector space. Now let $M_1^{\pm}(Z)$ be the space of signed probability measures. While it is not a subspace of $M^{\pm}(Z)$ as it does not include the zero measure, it is an affine subspace. Therefore, up to the choice of an arbitrary origin, it is a vector space.

Recall that an ensemble space must be a convex subset of a vector space. If we are able to represent ensembles as quasiprobability distributions over $Z$, then, $M_1^{\pm}(Z)$ will be isomorphic to the embedding vector space $V$ as topological vector spaces.\footnote{The question of whether such a $Z$ always exists is still open for us. A possible conjecture is that $Z$ could be constructed from the statistical quantities $F_i$ that fully identify each ensemble in $\Ens$.} In this case, any element $v\in V$ can be represented uniquely by a signed probability measure $\mu_{v}$, regardless of whether it is an actual ensemble or not. Since statistical quantities $F_i$ are affine functions, we will have $F_i(v) = \int_Z F_i(z) d\mu_{v}$. In other words, there is nothing special about quantum mechanics that allows quasiprobability representations. It is the fact that the ensemble space embeds into a vector space, a general fact for all theories, that allows such representations.

@@

Quasiprobability distributions are very useful in practice, particularly because one can construct them to fit particular problems. Once one has identified a set of observables that are the most natural ones to define a problem, one may find a representation that is specifically suited for the case. While the quasiprobability cannot be interpreted as probability, it still provides a useful tool to describe and understand the problem and give some intuition.

However, this practical convenience is also a drawback for finding a more general approach to probability in spaces beyond classical mechanics. Crucially, the space $Z$ over which the quasiprobability measures are defined is not unique, and it is not defined by the structure of the ensemble space and, as a consequence, cannot tell us which quasiprobability measures represent valid ensembles and which do not.

As a comparison, in the previous case all probability measures were given over a uniquely defined set of points, the extreme points of the ensemble space, but the probability measure may not be unique. In this case, the quasiprobability associated to each ensemble is unique, but the space over which the quasiprobability is defined is not.

@@

\textbf{Uniqueness of the fraction capacity.} From a conceptual standpoint, the fraction capacity has a major advantage: it has a clear physical meaning and it is unique. Like in Choquet theory and unlike for quasiprobability, the domain over which the measure is defined is fully determined by the system, by the ensemble space. On the other hand, unlike Choquet theory and like quasiprobability, once the space is fixed, there is one unique fraction capacity for each ensemble. Moreover, the setup is more general than Choquet theory (i.e. the ensemble space need not be compact) and stems from necessary physical requirements. An in depth study on fraction capacities, then, is guaranteed to be applicable to all physical theories, including those yet to be formulated.
